Uncovering underlying high-level musical content in the temporal domain \\ \small Leveraging deep learning, inductive bias, and aural music skills to learn embedding spaces from raw waveforms with applications to MIR downstream tasks

### 2023-05-09 methods/subsections/experiment setup and hyperparameter tuning.tex
Updates from Overleaf

For finetuning and improving the model's behavior, we followed \href{http://karpathy.github.io/2019/04/25/recipe/}{\textit{A Recipe for Training Neural Networks}} \cite{AndrejKarpathyHttp://karpathy.github.io/2019/04/25/recipe/} by Andrej Karpathy. A blog post that discusses a recipe for training neural networks. The author of the blog aims to expand on common mistakes people make when training neural nets by explaining how to avoid them through a particular process. Andrej Karpathy notes that there is often a gap between understanding the technical aspects of a neural network and achieving state-of-the-art results.

>>>>

For finetuning and improving the model's behavior, we followed \textit{A Recipe for Training Neural Networks} \cite{AndrejKarpathyHttp://karpathy.github.io/2019/04/25/recipe/} by Andrej Karpathy. A blog post that discusses a recipe for training neural networks. The blog aims to expand on common mistakes people make when training neural nets by explaining how to avoid them through a particular process. Andrej Karpathy notes that there is often a gap between understanding the technical aspects of a neural network and achieving state-of-the-art results.

### 2023-05-09 methods/subsections/model architecture and training strategy.tex
Updates from Overleaf

Self-supervised learning is a subfield of machine learning where models learn to generate representations of data by finding patterns and structures within the data without relying on external labels or annotations. In other words, the model learns by creating its supervision signal from the raw, unlabeled data.\cite{audioselfsupsurvey}

This approach to learning is inspired by the way humans and animals learn from their environment, where a great deal of knowledge is acquired through observation and interaction without explicit instruction. Self-supervised learning algorithms often involve tasks where the model learns to predict or reconstruct parts of the input data, such as predicting the next word in a sentence or completing an image with missing pixels.

Some of the main advantages of self-supervised learning include:

@@

In pattern recognition, invariance refers to the property of a model or algorithm being robust to certain transformations or changes in the input data. In other words, an invariant model should produce the same or similar output for different instances of the same object, even if they have undergone some changes.

@@

Unknown invariance: Unknown invariances are transformations or change the model has not been explicitly designed to handle. However, it still manages to be robust due to its learning capabilities. These invariances can emerge from the data and may not be known a priori. For example, a deep learning model might automatically learn to be invariant to certain lighting conditions or other factors not explicitly specified in its design.

@@

\begin{enumerate}
\item \textbf{Data Augmentation:} Create different augmentations or transformations of the same musical piece as positive pairs and use other randomly selected pieces as antagonistic pairs.
\item \textbf{Contrastive Learning:} Train by minimizing a contrastive loss function, encouraging the model to produce similar positive and dissimilar representations for antagonistic pairs.
\item\textbf{Latent Representation:} Utilize the learned representations for various MIR downstream tasks, such as music genre classification or similarity, both in the symbolic and time domains.
\end{enumerate}

\subsection{Siamese Networks}
\subsection{Triplet Siamese Networks}
\subsection{XXX}
\subsection{XXX}

@@

\subsection{Audio augmentation and transformation pipeline}
\subsubsection{Positive sample generation}
Let $W$ be the input waveform and $W_p$ be the generated positive waveform. The pipeline applies audio effects and adds noise with a random SNR. Define the following random variables for the effects:

\begin{equation}
gain \sim \text{Uniform}(-12, 0)
\end{equation}
\begin{equation}
pitch \sim \text{Uniform}(-1200, 1200)
\end{equation}
\begin{equation}
reverb\_params \sim \text{Uniform}(0, 100)^3
\end{equation}
\begin{equation}
chorus\_params = \begin{cases}
\text{Uniform}(0.1, 1.0), & i=1,2, \\
\text{Uniform}(20, 55), & i=3, \\
\text{Uniform}(0.1, 0.9), & i=4, \\
\text{Uniform}(0.1, 2.0), & i=5, \\
\text{Uniform}(2, 5), & i=6, \\
\text{Categorical}(\{-s, -t\}), & i=7,
\end{cases}
\end{equation}
\begin{equation}
drive \sim \text{Uniform}(0, 30)
\end{equation}
\begin{equation}
stretch \sim \text{Uniform}(0.9, 1.1)
\end{equation}
\begin{equation}
speed \sim \text{Uniform}(0.9, 1.1)
\end{equation}
\begin{equation}
tremolo\_speed \sim \text{Uniform}(0.1, 100)
\end{equation}
\begin{equation}
tremolo\_depth \sim \text{Uniform}(1, 101)
\end{equation}
\begin{equation}
snr\_range \sim \text{Uniform}(12, 100)
\end{equation}

Let $N$ be the noise matrix with dimensions [channels, time domain samples] as $W$ for white noise. Calculate signal power $P_s$ and noise power $P_n$:

\begin{equation}
P_s = \sum (W_{ij})^2
\end{equation}
\begin{equation}
P_n = \sum (N_{ij})^2
\end{equation}

Calculate the scaling factor $\alpha$ for noise based on the desired signal-to-noise ratio (SNR):

\begin{equation}
\alpha = \sqrt{\frac{P_s}{P_n \times 10^{\frac{snr\_range}{10}}}}
\end{equation}

Compute the scaled noise matrix $N_s$ and the noisy waveform matrix $W_n$:

\begin{equation}
N_s = \alpha \times N
\end{equation}
\begin{equation}
W_n = W + N_s
\end{equation}

The final positive waveform $W_p$ is obtained with noise added according to the chosen SNR. This pipeline enhances the diversity of the positive samples and aims to improve the model's ability to retrieve content-based similar-sounding samples sharing the same underlying musical language.

@@

Let $a$ be the original anchor clip and $n$ be the negative clip generated. The steps for generating the negative clip $n$ are as follows:

\begin{enumerate}
\item Calculate the length $L_a$ and duration $T_a$ of the anchor clip $a$:

\begin{equation}
L_a = \text{len}(a)
\end{equation}

\begin{equation}
T_a = \frac{L_a}{f_s}
\end{equation}

where $\text{len}(a)$ returns the shape of the audio clip, and $f_s$ is the sample rate of the audio.

\item Determine the number of chunks $n_c$ to split the anchor clip into based on a minimum chunk duration $T_c$:

\begin{equation}
n_c = \left\lfloor\frac{T_a}{T_c}\right\rfloor
\end{equation}

\item Calculate the minimum and maximum chunk lengths in samples:

\begin{equation}
L_{c,\text{min}} = T_c \times f_s
\end{equation}
\begin{equation}
L_{c,\text{max}} = T_{\text{max}} \times f_s
\end{equation}

where $T_{\text{max}}$ is the maximum chunk duration specified in seconds.

\item Generate $n_c$ chunk lengths $L_{c,1},\ldots,L_{c,n_c}$ as follows:

\begin{equation}
L_{c,i} = \begin{cases}
L_{c,\text{min}}, & i=1, \\
\text{randint}(L_{c,\text{min}}, L_{c,\text{max}}), & 1<i<n_c, \\
L_a - \sum_{j=1}^{i-1} L_{c,j}, & i=n_c.
\end{cases}
\end{equation}

\item Split the anchor clip into chunks $c_1,\ldots,c_{n_c}$ based on the lengths $L_{c,1},\ldots,L_{c,n_c}$:

\begin{equation}
c_i = a[..., s_i : s_i + L_{c,i}]
\end{equation}
\begin{equation}
S_c = [0] + [L_{c,1},...,L_{c,n_c-1}]
S_c = \sum_{i=0}^{n_c-1} S_{c,i}
\end{equation}
\begin{equation}
s_i = S_c[i]
\end{equation}
\begin{equation}
i=1,\ldots,n_c
\end{equation}

\item Shuffle the resulting list of chunks randomly to obtain the reordered chips $c'_1,\ldots,c'_{n_c}$:

Let $p$ be a random permutation of the indices $1$ to $n_c$. Then, the shuffled array $c'$ is given by:
\begin{equation}
c' = [c_{p(1)}, c_{p(2)}, \dots, c_{p(n_c)}]
\end{equation}

\item Let $c'_1, \dots, c'_{n_c}$ be the shuffled chunks. Then, the negative clip $n$ is obtained by concatenating the shuffled chunks:
\begin{equation}
n = [c'_1, \dots, c'_{n_c}]
\end{equation}

\end{enumerate}

@@

Triplet loss was initially proposed and used by Google to learn face recognition of the same person at different poses and angles. \cite{Schroff2015FaceNet:Clustering}

The mathematical expression for the triplet loss function with a variable margin can be written as follows:

\begin{equation}
L(\mathbf{a}, \mathbf{p}, \mathbf{n}) = \frac{1}{N} \sum_{i=1}^{N} \max \left(0, m + \left| \mathbf{a}_i - \mathbf{p}_i \right|_2^2 - \left| \mathbf{a}_i - \mathbf{n}_i \right|_2^2 \right)
\end{equation}

In the triplet loss function, $d(\mathbf{a}, \mathbf{p})$ and $d(\mathbf{a}, \mathbf{n})$ denote the Euclidean distance (L2-norm) between the anchor $\mathbf{a}$ and the positive sample $\mathbf{p}$ and the negative sample $\mathbf{n}$, respectively. The margin $m$ is a positive constant determining the minimum distance between the anchor-positive and anchor-negative pairs.

The $\max(0, ...)$ function computes the maximum of the expression within the parentheses and 0. This function ensures that the loss is only calculated for triplets that violate the margin constraint, where the negative sample is closer to the anchor than the positive sample by at least the margin. The loss is zero when the distance between the positive and negative samples exceeds the margin. Otherwise, the loss is proportional to the extent to which the constraint is violated.

@@

\begin{equation}
L = \frac{1}{N} \sum_{i=1}^{N} L(anchor_i, positive_i, negative_i)
\end{equation}

>>>>

Self-supervised learning (SSL) is a subfield of machine learning where models learn to generate representations of data by finding patterns and structures within the data without relying on external labels or annotations. In other words, the model learns by creating its supervision signal from the raw, unlabeled data.\cite{audioselfsupsurvey}

This approach to learning is inspired by the way humans and animals learn from their environment, where a great deal of knowledge is acquired through observation and interaction without explicit instruction. SSL algorithms often involve tasks where the model learns to predict or reconstruct parts of the input data, such as predicting the next word in a sentence or completing an image with missing pixels.

Some of the main advantages of SSL include:

@@

In this section, we discuss the concepts of known and unknown invariance in pattern recognition. Invariance refers to the property of a model or algorithm being robust to certain transformations or changes in the input data. In other words, an invariant model should produce the same or similar output for different instances of the same object, even if they have undergone some changes.

@@

Unknown invariance: Unknown invariances are transformations or changes the model has not been explicitly designed to handle. However, it still manages to be robust due to its learning capabilities. These invariances can emerge from the data and may not be known a priori. For example, a deep learning model might automatically learn to be invariant to certain lighting conditions or other factors not explicitly specified in its design.

@@

\begin{enumerate}
\item \textbf{Data Augmentation:} Create different augmentations or transformations of the same musical piece as positive pairs and use other randomly selected pieces as antagonistic pairs.
\item \textbf{Contrastive Learning:} Train by minimizing a contrastive loss function, encouraging the model to produce similar representations for positive pairs and dissimilar representations for antagonistic pairs.
\item\textbf{Latent Representation:} Utilize the learned representations for various Music Information Retrieval (MIR) downstream tasks, such as music genre classification or similarity, both in the symbolic and time domains.
\end{enumerate}

By following these steps, CLMR models can learn meaningful and robust representations of music, potentially enhancing their performance in various MIR tasks and reducing the need for labeled data.

\subsection{Siamese Networks}

A Siamese Network, a deep learning architecture introduced by Bromley and LeCun in the early 1990s \cite{Bromley1993SignatureNetwork}, is designed explicitly for tasks involving similarity or comparison between two input instances, such as signature verification. Its name derives from the structure, which consists of two or more identical subnetworks connected in parallel and joined at the output layer. These subnetworks share the same architecture, weights, and hyperparameters, allowing for more efficient memory usage and computational complexity. This design is particularly effective for learning from limited or imbalanced datasets, as it focuses on learning the similarity metric rather than specific features of individual classes.

Each subnetwork processes an input independently, and its outputs are combined and further processed to yield a single similarity score. The shared weights during training enable the model to learn an invariant representation for the input instances, improving its efficiency in comparing and contrasting them. Siamese Networks employ specialized loss functions, such as contrastive or triplet loss, which minimize the distance between similar input pairs and maximize the distance between dissimilar ones. This approach encourages the network to learn a meaningful similarity metric.

\subsection{Triplet Siamese Networks}

Triplet Siamese Networks, an extension of the Siamese Network architecture, involve comparing three input instances instead of two, aiming to learn an embedding space where similar instances are close and dissimilar instances are distant. The input consists of an anchor, a positive instance (same class as the anchor), and a negative instance (different class). Each instance is processed by an identical subnetwork with shared architecture, weights, and hyperparameters.

The triplet loss function guides learning, minimizing the distance between the anchor and positive instances while maximizing the distance between the anchor and negative instances. A margin parameter in the loss function ensures a minimum separation between positive and negative instances in the embedding space.

Triplet Siamese Networks offer several advantages, including learning fine-grained similarity metrics, being well-suited for one-shot learning tasks, and offering efficient memory usage and computational complexity due to shared weights. These networks have been applied to various tasks and domains, such as image recognition, face recognition, speaker recognition, text similarity, and medical imaging for diagnosis or treatment purposes.

\subsection{XXX}
\subsection{XXX}

@@

\subsection{Audio augmentation and transformation pipeline}

%\input{figures/neural networks/pipeline}

\subsubsection{Positive sample generation}
Given an anchor audio signal $A[n]$, we generate a positive signal $P[n]$ by applying a series of audio effects and adding noise. Let $g \in [-12, 0]$ represent the gain, $\Delta p \in [-1200, 1200]$ the pitch shift, $h_R[n]$ the impulse response of the reverb with parameters in $[0, 100]$, $h_C[n]$ the impulse response of the chorus with parameters determined by the specified ranges, $d \in [0, 30]$ the overdrive parameter, $\alpha \in [0.9, 1.1]$ the speed change factor, $\beta \in [0.9, 1.1]$ the stretch factor, $f_m \in [0.1, 100]$ the modulation frequency of the tremolo, and $d_m \in [1, 101]$ the depth of the tremolo.

The positive signal $P[n]$ is generated as follows:

\begin{equation}\label{eq:positive_signal}
P[n] = A[n] \ast h_{G}[n] \ast h_{C}[n] \ast h_{D}[n] \ast h_{P_t}[n] \ast h_{R}[n] \ast h_{S}[n] \ast h_{T}[n] \ast h_{T_m}[n] + N[n]
\end{equation}

where $\ast$ denotes convolution, $h_{G}[n] = g \delta[n]$, $h_{D}[n]$ is the impulse response of the overdrive effect, $h_{P_t}[n]$ is the impulse response of the pitch shift effect, $h_{S}[n]$ represents the impulse response of the speed change effect, $h_{T}[n]$ is the impulse response of the stretch effect, and $h_{T_m}[n] = (1 - d_m \cos(2 \pi f_m n))\delta[n]$ is the impulse response of the tremolo effect. The noise signal $N[n]$ is added with a specific signal-to-noise ratio (SNR) in the range $[12, 100]$ in decibels.

These effect selections behave and affect the signal in several domains, adding aggressive transformations while preserving high-level musical content.

\begin{itemize}
\item \textbf{Amplitude effects:} Amplitude or gain modification.
\begin{itemize}
    \item Gain: Adjusts the overall amplitude of the signal by a constant factor $g \in [-12, 0]$ dB.
\end{itemize}

\item \textbf{Time-domain effects:} Time-domain manipulation alters the signal's timing or duration.
\begin{itemize}
    \item Speed change: Alters the playback speed of the signal by a factor $\alpha \in [0.9, 1.1]$.
    \item Stretch: Changes the signal duration without affecting its pitch by a factor $\beta \in [0.9, 1.1]$.
\end{itemize}

\item \textbf{Frequency-domain effects:} Frequency domain manipulation alters the frequency content or pitch of the signal.
\begin{itemize}
    \item Pitch shift: Changes the signal's pitch by $\Delta p \in [-1200, 1200]$ cents.
\end{itemize}

\item \textbf{Nonlinear effects:} Nonlinear transformations to the audio signal, resulting in harmonic distortion or other complex changes.
\begin{itemize}
    \item Overdrive: Adds harmonic distortion to the signal based on the parameter $d \in [0, 30]$.
\end{itemize}

\item \textbf{Modulation effects:} Audio signal modulation using a low-frequency oscillator or control signal.
\begin{itemize}
    \item Chorus: Applies a varying time delay to the signal, resulting in a richer, thicker sound. The specified ranges determine the chorus parameters.
    \item Tremolo: Modulates the signal's amplitude at a modulation frequency $f_m \in [0.1, 100]$ Hz and depth $d_m \in [1, 101]$.
\end{itemize}

\item \textbf{Reverberation effects:} Acoustic reflections simulation and reverberations of physical space.
\begin{itemize}
    \item Reverb: Applies an impulse response $h_R[n]$ that simulates the reverberation of space with parameters in the range $[0, 100]$.
\end{itemize}

\item \textbf{Noise effects:} These effects add noise or other random elements to the audio signal.
\begin{itemize}
    \item Additive noise: Adds a noise signal $N[n]$ with a specific signal-to-noise ratio (SNR) in the range $[12, 100]$.
\end{itemize}
\end{itemize}

@@

\begin{enumerate}
\item Calculate the minimum and maximum chunk lengths in samples:
2\begin{equation}
l_{min} = t_{min} \times S
\end{equation}
\begin{equation}
l_{max} = t_{max} \times S
\end{equation}
\item Generate random chunk lengths $l_1, l_2, \ldots, l_{n-1}$ from the uniform distribution on the interval $[l_{min}, l_{max}]$. Calculate the final chunk length as:
\begin{equation}
l_n = L_A - \sum_{i=1}^{n-1} l_i
\end{equation}
where $L_A$ is the length of the anchor signal in samples.
\item Split the anchor signal $A$ into chunks $C_1, C_2, \ldots, C_n$ according to the calculated chunk lengths.
\item Shuffle the chunks randomly to get the permuted chunks $C_{\sigma(1)}, C_{\sigma(2)}, \ldots, C_{\sigma(n)}$, where $\sigma$ is a random permutation of indices from $1$ to $n$.
\item Concatenate the shuffled chunks to generate the negative signal:
\begin{equation}\label{eq:negative_signal}
N = C_{\sigma(1)} \oplus C_{\sigma(2)} \oplus \ldots \oplus C_{\sigma(n)}
\end{equation}
\end{enumerate}

Ensure that the length of the anchor and negative signals are the same, i.e.,

\begin{equation}
L_A = L_N,
\end{equation}

where $L_N$ is the length of the negative signal in samples.

@@

Schroff, F., Kalenichenko, D. and Philbin, J. at Google initially proposed and used triplet loss to learn face recognition of the same person at different poses and angles. \cite{Schroff2015FaceNet:Clustering}

A triplet loss function with a variable margin can be written as follows:

\begin{equation}
\mathcal{L}(\mathbf{a}, \mathbf{p}, \mathbf{n}) = \sum_{i=1}^{N} \max \left(0, \left| \mathbf{a}_i - \mathbf{p}_i \right|_2^2 - \left| \mathbf{a}_i - \mathbf{n}_i \right|_2^2 + m \right)
\end{equation}

In this equation:

$\mathcal{L}$ is the triplet loss function.
$\mathbf{a}_i$, $\mathbf{p}_i$, and $\mathbf{n}_i$ are the anchor, positive, and negative embedding vectors, respectively, for the $i$-th sample.
$\left| \cdot \right|_2^2$ denotes the squared Euclidean distance between two points.
$m$ is the margin, a hyperparameter that helps ensure that the anchor-positive distance is smaller than the anchor-negative distance by at least a margin.
The summation $\sum_{i=1}^{N}$ is over all $N$ triplets in the dataset.
The goal of the triplet loss is to minimize the distance between the anchor and the positive sample while maximizing the distance between the anchor and the negative sample. This encourages the model to learn embeddings that are similar for the same class and dissimilar for different classes.

@@

\begin{equation}
\mathcal{L} = \frac{1}{N} \sum_{i=1}^{N} L(anchor_i, positive_i, negative_i)
\end{equation}
